### 2018-06-12 express2018/formal.tex
text
Fixed formal.tex and sos.tex (releasing locks on my side)

We highlight here a formalization of CCS
in the HOL theorem
prover (HOL4) \cite{slind2008brief},
including the new concepts and theorems proposed in the first half of
this paper.
%  The main purpose is to convince the readers that, there's no flaw
%  in the informal proofs. 
The whole formalisation can be found 
in \cite{Tian:2017wrba}  or in 
 in the official example folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.
\finish{i would say here how big is the formalisation, ie how many
  lines of code} 
%  This
%  work is based on a formalization of CCS in HOL theorem
% prover (HOL4) \cite{slind2008brief}, described in second author's MSc thesis
% \cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
% code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.

Higher Order Logic (or ``HOL Logic'') \cite{hollogic} stands for simple-typed $\lambda$-calculus plus Hibert
choice operator, axiom of infinity, and rank-1 polymorphim (type
variables). HOL4 implemented this logic without any extension, in
contrast
with
some other theorem provers of the  HOL family (e.g. Isabelle/HOL). 
\finish{not having extensions above: sounds like a defect; what is the
advantage? the portability explained in the next sentence?} 
The HOL Logic has considerability simpler logic
foundations than most other theorem provers, e.g. Coq. As a result,
formal theories in HOL can be easily migrated (sometimes even
automatically) into other theorem provers.

\finish{below: to be reformulated} 
We will not repeat all definitions and theorems of CCS again by their
formalized versions. Instead, we just focus on several highlights in
this work, i.e.
\begin{enumerate}
\item The formalization of semantics context in $\lambda$-expressions and the theory of
  (pre)congruence for CCS;
\item The definition and uses of trace in the proof of unique solution of
  contractions theorem;
\item The proof of the hard version of ``coarsest congruence
  contained in $\approx$'' theorem.
\end{enumerate}

In contrast with the theorems in the first part of the paper, in the
formalisation we consider only single equations/contractions. 
This considerably simplifies the  reading of the HOL scripts. 
The extension of the work to multiple equations/contractions is
discussed in Section~\ref{s:}.

>>>>

We highlight here a formalization of CCS
in the HOL theorem
prover (HOL4) \cite{slind2008brief},
including the new concepts and theorems proposed in the first half of
this paper.
%  The main purpose is to convince the readers that, there's no flaw
%  in the informal proofs. 
The whole formalization can be found 
in \cite{Tian:2017wrba}  or in 
 in the official example folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}. The
work consists of about 20,000 lines of proof scripts in Standard ML.
%  This
%  work is based on a formalization of CCS in HOL theorem
% prover (HOL4) \cite{slind2008brief}, described in second author's MSc thesis
% \cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
% code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.

Higher Order Logic (or ``HOL Logic'') \cite{hollogic} stands for simple-typed $\lambda$-calculus plus Hibert
choice operator, axiom of infinity, and rank-1 polymorphim (type
variables). HOL4 implemented the original HOL Logic, while
some other theorem provers of the HOL family (e.g. Isabelle/HOL) added
further extensions (they made the formal language more powerful,
but they also bring the possibilities that the entire logic becomes inconsistent).
The HOL Logic has considerability simpler logic
foundations than most other theorem provers, e.g. Coq. As a result,
formal theories in HOL can be easily migrated (sometimes even
automatically) into other theorem provers.

\finish{below: to be reformulated (by Davide) later} 
We will not repeat all definitions and theorems of CCS again by their
formalized versions. Instead, we just focus on several highlights in
this work, i.e.
\begin{enumerate}
\item The formalization of semantics context in $\lambda$-expressions and the theory of
  (pre)congruence for CCS;
\item The definition and uses of trace in the proof of unique solution of
  contractions theorem;
\item The proof of the hard version of ``coarsest congruence
  contained in $\approx$'' theorem.
\end{enumerate}

In contrast with the theorems in the first part of the paper, in the
formalization we consider only single equations/contractions. 
This considerably simplifies the  reading of the HOL scripts. 
The extension of the work to multiple equations/contractions is
discussed in Section~\ref{s:}. \finish{to be finished}

### 2018-06-12 express2018/sos.tex
Fixed formal.tex and sos.tex (releasing locks on my side)

In the formalization of CCS, it's better to have two distinct
types for actions: the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Label}} ($\beta$ is a type variable.) for visible actions, divided by input
and output actions:
\begin{lstlisting}
val _ = Datatype `Label = name 'b | coname 'b`;
\end{lstlisting}
and the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Action}} as the
union of all visible and invisible actions. The cardinality of all
available types depends on the choice of type variable $\beta$.

The core datatype ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS}}'' is then defined as an inductive
datatype in HOL, based on its Datatype package:
\begin{lstlisting}
val _ = Datatype `CCS = nil
		      | var 'a
		      | prefix ('b Action) CCS
		      | sum CCS CCS
		      | par CCS CCS
		      | restr (('b Label) set) CCS
		      | relab CCS ('b Relabeling)
		      | rec 'a CCS`;
\end{lstlisting}

We have added some minimal grammar support using HOL's powerful pretty printer, to represent CCS
processes in more readable forms. Table \ref{tab:ccsoperator} has listed the notation of typical CCS processes and
major operators supported by above definition. (Notice the use of
recursition operator for representing process constants)
\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Operator} & \textbf{Notation} & \textbf{HOL term} & \textbf{HOL
  (alternative)}\\
\hline
nil & $\textbf{0}$ &  \HOLinline{\HOLConst{nil}} & \HOLinline{\HOLConst{nil}} \\
Prefix & $a.b.0$ & \texttt{prefix a (prefix b nil)} & \HOLinline{\HOLFreeVar{a}\HOLSymConst{..}\HOLFreeVar{b}\HOLSymConst{..}\HOLConst{nil}} \\
Sum & $p + q$ & \texttt{sum p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{+} \HOLFreeVar{q}} \\
Parallel & $p \,\mid\, q$ & \texttt{par p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{q}} \\
Restriction & $(\nu\;L)\;p$ & \texttt{restr L p} & \HOLinline{\HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{p}} \\
Constant & $A=a.A$ & \texttt{rec A (prefix a (var A))} & \HOLinline{\HOLConst{rec} \HOLFreeVar{A} (\HOLFreeVar{a}\HOLSymConst{..}\HOLConst{var} \HOLFreeVar{A})} \\
\hline
\end{tabular}
\end{center}
\vspace{-1em}
   \caption{Syntax of CCS operators}
   \label{tab:ccsoperator}
\end{table}

The transition semantics of CCS processes were defined by the following
Structural Operational Semantics (SOS for short) rules in
Fig.\;\ref{f:LTSCCS}, except that the transition rule for constants
has been changed to recursion operator:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}) \HOLFreeVar{X} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}
\end{alltt}
which says, if we substitute all appearences of constant $X$ in $E$ to
$(\mathbf{rec}\; X\; E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; X\; E)$ has the same
transition. Other SOS rules have the following forms:
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{u}\HOLSymConst{..}\HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E}\hfill\texttt{[PREFIX]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E} \HOLSymConst{+} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[SUM1]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{+} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[SUM2]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sp{\prime}}\hfill\texttt{[PAR1]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[PAR2]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLConst{label} \HOLFreeVar{l}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLConst{label} (\HOLConst{COMPL} \HOLFreeVar{l})\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{E} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLSymConst{\ensuremath{\tau}}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sb{\mathrm{2}}}\hfill\texttt{[PAR3]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} ((\HOLFreeVar{u} \HOLSymConst{=} \HOLSymConst{\ensuremath{\tau}}) \HOLSymConst{\HOLTokenDisj{}} (\HOLFreeVar{u} \HOLSymConst{=} \HOLConst{label} \HOLFreeVar{l}) \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{l} \HOLSymConst{\HOLTokenNotIn{}} \HOLFreeVar{L} \HOLSymConst{\HOLTokenConj{}} \HOLConst{COMPL} \HOLFreeVar{l} \HOLSymConst{\HOLTokenNotIn{}} \HOLFreeVar{L}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{E\sp{\prime}}\hfill\texttt{[RESTR]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{relab} \HOLFreeVar{E} \HOLFreeVar{rf} \HOLTokenTransBegin\HOLConst{relabel} \HOLFreeVar{rf} \HOLFreeVar{u}\HOLTokenTransEnd \HOLConst{relab} \HOLFreeVar{E\sp{\prime}} \HOLFreeVar{rf}\hfill\texttt{[RELABELING]}
\end{alltt}
Above rules work as \emph{axioms} in the theory of CCS, as they cannot be
proved by more primitive theorems or axioms. From the view of theorem prover, these
SOS rules are nothing but an \emph{inductive 
  definition} on 3-ary relation \HOLinline{\HOLConst{TRANS}} of type ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\beta} \HOLTyOp{Action} \HOLTokenTransEnd (\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \HOLTyOp{bool}}'', generated by HOL4's function
\texttt{Hol_reln}.\footnote{On the other side, above literal SOS rules are not
the \emph{only} thing we got from \texttt{Hol_reln}, which also
returns the fix-point theorem and induction principle for the
relation, otherwise things like ``$\textbf{0}$ has no transition''
cannot be proved at all.} The ultimate result here (before
bisimilution gets involved), is an ML function taking a CCS process
and returning a theorem indicating all its direct transitions.

For instance, we know that the process $(a.0 | \bar{a}.0)$ have three
possible transitions:
\begin{enumerate}
\item $(a.0 | \bar{a}.0) \overset{a}{\longrightarrow} (0 | \bar{a}.0)$;
\item $(a.0 | \bar{a}.0) \overset{\bar{a}}{\longrightarrow} (a.0 | 0)$;
\item $(a.0 | \bar{a}.0) \overset{\tau}{\longrightarrow} (0 | 0)$.
\end{enumerate}
To completely decide all possible transitions, if done manually, the
following results must be proved:
\begin{enumerate}
\item there exists transitions from $(a.0 | \bar{a}.0)$ (optionally);
\item each of above three transitions using SOS inference rules;
\item there's no other transition beside the three existing ones.
\end{enumerate}
For large complex processes it's surprising hard to prove the
non-existence of transitions. Instead, if we call the function \texttt{CCS\_TRANS\_CONV} with the
root process:
\begin{lstlisting}
> CCS_TRANS_CONV
	 ``par (prefix (label (name "a")) nil)
	       (prefix (label (coname "a")) nil)``
\end{lstlisting}
we get the following theorem, which completely answers the question:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{In} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{Out} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E} \HOLSymConst{\HOLTokenEquiv{}}
   ((\HOLFreeVar{u} \HOLSymConst{=} \HOLConst{In} \HOLStringLit{a}) \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{E} \HOLSymConst{=} \HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{Out} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil}) \HOLSymConst{\HOLTokenDisj{}}
    (\HOLFreeVar{u} \HOLSymConst{=} \HOLConst{Out} \HOLStringLit{a}) \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{E} \HOLSymConst{=} \HOLConst{In} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{nil})) \HOLSymConst{\HOLTokenDisj{}}
   (\HOLFreeVar{u} \HOLSymConst{=} \HOLSymConst{\ensuremath{\tau}}) \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{E} \HOLSymConst{=} \HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{nil})
\end{alltt}

>>>>

In the formalization of CCS, 
the type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' ($\beta$ is a
type variable) accounts for visible actions, divided into input
and output actions:
\begin{lstlisting}
val _ = Datatype `Label = name 'b | coname 'b`;
\end{lstlisting}
Then
the type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Action}}'' is the
union of all visible and invisible actions. The cardinality of
``\HOLinline{\ensuremath{\beta} \HOLTyOp{Action}}'' (and therefore of all
CCS types built on top of it) 
 depends on the choice of initial  type $\beta$.

The core datatype ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS}}'', accounting for the CCS syntax, is then defined as an inductive
datatype in HOL, based on its Datatype package: ($\alpha$ is the type
variable for recursion variables) 
\begin{lstlisting}
val _ = Datatype `CCS = nil
		      | var 'a
		      | prefix ('b Action) CCS
		      | sum CCS CCS
		      | par CCS CCS
		      | restr (('b Label) set) CCS
		      | relab CCS ('b Relabeling)
		      | rec 'a CCS`;
\end{lstlisting}

We have added some  grammar support using HOL's powerful pretty printer, to represent CCS
processes in more readable forms (as shown in the ``HOL (abbrev.)''
column in Table \ref{tab:ccsoperator}, the rest of table has listed the notation of typical CCS processes and
major operators supported by above definition). For restriction
operator, we chose to takes a set of names, rather than a
  single name as in the  CCS syntax, to prevent the formalization
  efforts on the equivalence of CCS processes with different order of
  nested restrictions. Also, we don't assume that al recursion are
  guarded, i.e. any term like \texttt{var X} must appear in \texttt{P} of \texttt{rec X P}.

%  (Notice the use of
% recursition operator for representing process constants)
\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Operator} & \textbf{CCS Notation} & \textbf{HOL term} &
                                                                \textbf{HOL (abbrev.)}\\
\hline
nil & $\textbf{0}$ & \HOLinline{\HOLConst{nil}} & \HOLinline{\HOLConst{nil}} \\
prefix & $u.P$ & \texttt{prefix u P} & \HOLinline{\HOLFreeVar{u}\HOLSymConst{..}\HOLFreeVar{P}} \\
sum & $P + Q$ & \texttt{sum P Q} & \HOLinline{\HOLFreeVar{P} \HOLSymConst{+} \HOLFreeVar{Q}} \\
parallel & $P \,\mid\, Q$ & \texttt{par P Q} & \HOLinline{\HOLFreeVar{P} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{Q}} \\
restriction & $(\nu\;L)\;P$ & \texttt{restr L P} & \HOLinline{\HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{P}}
  \\
recursion & $A$ with $A\Defi P$ & \texttt{rec A P} & \HOLinline{\HOLConst{rec} \HOLFreeVar{A} \HOLFreeVar{P}} \\
input action & $a$ & \texttt{label (name a)} & \HOLinline{\HOLConst{In} \HOLFreeVar{a}} \\
output action & $\outC a$ & \texttt{label (coname a)} & \HOLinline{\HOLConst{Out} \HOLFreeVar{a}} \\
\hline
\end{tabular}
\end{center}
\vspace{-1em}
   \caption{Syntax of CCS operators}
   \label{tab:ccsoperator}
\end{table}

The transition semantics of CCS processes  follows
the Structural Operational Semantics (SOS) rules in
Figure~\ref{f:LTSCCS}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{u}\HOLSymConst{..}\HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E}\hfill\texttt{[PREFIX]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E} \HOLSymConst{+} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[SUM1]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{+} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[SUM2]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sp{\prime}}\hfill\texttt{[PAR1]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[PAR2]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLConst{label} \HOLFreeVar{l}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLConst{label} (\HOLConst{COMPL} \HOLFreeVar{l})\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{E} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sp{\prime}} \HOLTokenTransBegin\HOLSymConst{\ensuremath{\tau}}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{E\sb{\mathrm{2}}}\hfill\texttt{[PAR3]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} ((\HOLFreeVar{u} \HOLSymConst{=} \HOLSymConst{\ensuremath{\tau}}) \HOLSymConst{\HOLTokenDisj{}} (\HOLFreeVar{u} \HOLSymConst{=} \HOLConst{label} \HOLFreeVar{l}) \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{l} \HOLSymConst{\HOLTokenNotIn{}} \HOLFreeVar{L} \HOLSymConst{\HOLTokenConj{}} \HOLConst{COMPL} \HOLFreeVar{l} \HOLSymConst{\HOLTokenNotIn{}} \HOLFreeVar{L}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{E\sp{\prime}}\hfill\texttt{[RESTR]}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{relab} \HOLFreeVar{E} \HOLFreeVar{rf} \HOLTokenTransBegin\HOLConst{relabel} \HOLFreeVar{rf} \HOLFreeVar{u}\HOLTokenTransEnd \HOLConst{relab} \HOLFreeVar{E\sp{\prime}} \HOLFreeVar{rf}\hfill\texttt{[RELABELING]}
\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}) \HOLFreeVar{X} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}\hfill\texttt{[REC]}
\end{alltt}

Here the rule \texttt{[REC]} (Recursion)
 says that if we substitute all appearences of variable $A$ in $E$ to
$(\mathbf{rec}\; A\; E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; A\; E)$ has the same
transition. From the HOL  viewpoint, these
SOS rules are \emph{inductive 
  definitions} on the 3-ary relation \HOLinline{\HOLConst{TRANS}} of type ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\beta} \HOLTyOp{Action} \HOLTokenTransEnd (\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \HOLTyOp{bool}}'', generated by HOL4's 
\texttt{Hol_reln} function.\footnote{\texttt{Hol_reln} also
gives us a fix-point theorem and induction principle for the
relation, otherwise statements like ``$\textbf{0}$ has no transition''
could not be proved.}

HOL4 is written in Standard ML (or ML for abbreviation). This single language plays three roles here:
\begin{enumerate}
\item It serves as the underlying implementation language for the core HOL engine;
\item It's used to implement tactics (and tacticals) for automatic theorem proving;
\item It's used as the command language of the interactive HOL system.
\end{enumerate}
Besides, HOL4 users can write arbitrary applications which leverages
the theorem proving facility. The ultimate result (initially done by Nesi)
at this point, is a complex Standard ML function
  taking a CCS process and returning a theorem indicating all its direct transitions.

For instance, we know that the process $(a.0 | \bar{a}.0)$ have three
possible transitions: $(a.0 | \bar{a}.0) \overset{a}{\longrightarrow}
(0 | \bar{a}.0)$, $(a.0 | \bar{a}.0)
\overset{\bar{a}}{\longrightarrow} (a.0 | 0)$ and $(a.0 | \bar{a}.0)
\overset{\tau}{\longrightarrow} (0 | 0)$.
To completely decide all possible transitions, if done manually, the
following facts should be proved:
\begin{enumerate}
\item there exists transitions from $(a.0 | \bar{a}.0)$ (optional);
\item the correctness for each of the transitions by SOS rules;
\item there's no other transition beside the three existing ones.
\end{enumerate}
For large complex processes it's surprisingly hard to manually prove the
non-existence of transitions. Instead, if we call the ML function \texttt{CCS\_TRANS\_CONV} on the
root process $(a.0 | \bar{a}.0)$:
\begin{lstlisting}
> CCS_TRANS_CONV
	 ``par (prefix (label (name "a")) nil)
	       (prefix (label (coname "a")) nil)``
\end{lstlisting}
We immediately get the following theorem, which completely answers the question:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{In} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{Out} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E} \HOLSymConst{\HOLTokenEquiv{}}
   ((\HOLFreeVar{u} \HOLSymConst{=} \HOLConst{In} \HOLStringLit{a}) \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{E} \HOLSymConst{=} \HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{Out} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil}) \HOLSymConst{\HOLTokenDisj{}}
    (\HOLFreeVar{u} \HOLSymConst{=} \HOLConst{Out} \HOLStringLit{a}) \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{E} \HOLSymConst{=} \HOLConst{In} \HOLStringLit{a}\HOLSymConst{..}\HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{nil})) \HOLSymConst{\HOLTokenDisj{}}
   (\HOLFreeVar{u} \HOLSymConst{=} \HOLSymConst{\ensuremath{\tau}}) \HOLSymConst{\HOLTokenConj{}} (\HOLFreeVar{E} \HOLSymConst{=} \HOLConst{nil} \HOLSymConst{\ensuremath{\parallel}} \HOLConst{nil})
\end{alltt}
